Type Classes for Mathematics in Type Theory

### 2010-06-28 jar.tex
In same paragraph, add remark about operational type classes.
Add paragraph about re-use of operational type classes. Fix some code whitespace formatting.

% todo:
  % cite that other type class based development in the context of the alternative bundling schemes.
  % maybe have a chapter called "interfaces are primary, not implementations" (also involve polynomials)
  % mention that we use same RingPlus/RingMult operational type classes (with notation) in SemiRing/Ring.

@@

Higher structures, too, can all be expressed as predicates (expressing laws) over a number of carriers, relations, and operations. While optimally flexible in principle, in practice \emph{naive} adoption of this approach (that is, without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest. And when \emph{applying} such theorems, one must either enumerate any non-inferrable components, or let the system spawn awkward metavariables to be resolved at a later time. Importantly, this also hinders proof search for proofs of the structure predicates, making any nontrivial use of theorems a laborious experience. Finally, the lack of \emph{canonical names} for particular components of abstract structures makes it impossible to associate idiomatic notations with them.\marginpar{something about opacity?}

@@

Because predicate classes only provide contextual information and are insulated from the actual algebraic expressions, their instances can always be kept entirely opaque---only their existence matters. Together, these properties largely defuse an argument occasionally voiced against type classes concerning perceived unpredictability of instance resolution. While it is certainly true that in contexts with redundant information it can become hard to predict which instance of a predicate class will be found by proof search, this is simply a non-issue because it \emph{does not matter} which one is found.

@@

@@

With these in place, our type class for categories follows the usual type-theoretical definition of a
category~\cite{saibi1995constructive}:
\begin{lstlisting}
Class Category (O: Type) `{!Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
    `{!CatId O} `{!CatComp O}: Prop :=
  { arrow_equiv:> Π x y, Setoid (x ⟶ y)
  ; comp_proper:> Π x y z, Proper (equiv ==> equiv ==> equiv) comp
  ; comp_assoc w x y z (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
      c ◎ (b ◎ a) = (c ◎ b) ◎ a
  ; id_l `(a: x ⟶ y): cat_id ◎ a = a
  ; id_r `(a: x ⟶ y): a ◎ cat_id = a }.
\end{lstlisting}
This definition is based on the 2-categorical idea of having equality only on arrows, not on objects.
%Similarly, we will have equality on natural transformations, but not on functors. % todo: ORLY?

@@

  Class Functor `(Fmap): Prop :=
    { functor_from: Category C
    ; functor_to: Category D
    ; functor_morphism:> Π a b: C, Setoid_Morphism (@fmap _ a b)
    ; preserves_id: `(fmap (cat_id: a ⟶ a) = cat_id)
    ; preserves_comp `(f: y ⟶ z) `(g: x ⟶ y):
      fmap (f ◎ g) = fmap f ◎ fmap g }.
\end{lstlisting}
We ought to say a few words about our use of \lstinline|fmap|. The usual mathematical notational convention for functor application is to use the name of the functor to refer to both its object map and its arrow map, relying on additional conventions regarding object/arrow names for disambiguation: ``\lstinline|F x|'' and ``\lstinline|F f|'' map an object and an arrow, respectively, because ``\lstinline|x|'' and ``\lstinline|f|'' are conventional names for objects and arrows, respectively.

>>>>

% todo:
  % cite that other type class based development in the context of the alternative bundling schemes.
  % maybe have a chapter called "interfaces are primary, not implementations" (also involve polynomials)

@@

Higher structures, too, can all be expressed as predicates (expressing laws) over a number of carriers, relations, and operations. While optimally flexible in principle, in practice \emph{naive} adoption of this approach (that is, without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest. And when \emph{applying} such theorems, one must either enumerate any non-inferrable components, or let the system spawn awkward metavariables to be resolved at a later time. Importantly, this also hinders proof search for proofs of the structure predicates, making any nontrivial use of theorems a laborious experience. Finally, the lack of \emph{canonical names} for particular components of abstract structures makes it impossible to associate idiomatic notations with them.

@@

Because predicate classes only provide contextual information and are insulated from the actual algebraic expressions, their instances can always be kept entirely opaque---only their existence matters. Together, these properties largely defuse an argument occasionally voiced against type classes concerning perceived unpredictability of instance resolution. While it is certainly true that in contexts with redundant information it can become hard to predict which instance of a predicate class will be found by proof search, this is simply a non-issue because it \emph{does not matter} which one is found. And for operational type classes the issue rarely arises because their instances are not nearly as abundant.

@@

The operational type classes used in \lstinline|SemiRing| for zero, one, multiplication and addition, are the same ones used by \lstinline|Ring| and \lstinline|Field| (not shown). Thus, the realization that a particular semiring is in fact a ring or field (etc.) has no bearing on how one refers to the operations in question, which is as it should be. The realization that a particular semigroup is part of a semiring \emph{does} call for a new (canonical) name, simply because of the need for disambiguation. The introduction of these additional names for the same operation is quite harmless in practice, because canonical names established by operational type class fields are just trivial identity functions anyway, so that in most contexts the distinction reduces away instantly.

@@

With these in place, our type class for categories follows the usual type-theoretical definition of a
category~\cite{saibi1995constructive}:
\begin{lstlisting}
Class Category (O: Type) `{!Arrows O} `{Π x y: O, Equiv (x ⟶ y)}
    `{!CatId O} `{!CatComp O}: Prop :=
  { arrow_equiv:> Π x y, Setoid (x ⟶ y)
  ; comp_proper:> Π x y z, Proper (equiv ==> equiv ==> equiv) comp
  ; comp_assoc w x y z (a: w ⟶ x) (b: x ⟶ y) (c: y ⟶ z):
      c ◎$$ (b ◎$$ a) = (c ◎$$ b) ◎$$ a
  ; id_l `(a: x ⟶ y): cat_id ◎$$ a = a
  ; id_r `(a: x ⟶ y): a ◎$$ cat_id = a }.
\end{lstlisting}
This definition is based on the 2-categorical idea of having equality only on arrows, not on objects.
%Similarly, we will have equality on natural transformations, but not on functors. % todo: ORLY?

@@

  Class Functor `(Fmap): Prop :=
    { functor_from: Category C
    ; functor_to: Category D
    ; functor_morphism:> Π a b: C, Setoid_Morphism (@fmap _ a b)
    ; preserves_id: `(fmap (cat_id: a ⟶ a) = cat_id)
    ; preserves_comp `(f: y ⟶ z) `(g: x ⟶ y):
      fmap (f ◎$$ g) = fmap f ◎$$ fmap g }.
\end{lstlisting}
We ought to say a few words about our use of \lstinline|fmap|. The usual mathematical notational convention for functor application is to use the name of the functor to refer to both its object map and its arrow map, relying on additional conventions regarding object/arrow names for disambiguation: ``\lstinline|F x|'' and ``\lstinline|F f|'' map an object and an arrow, respectively, because ``\lstinline|x|'' and ``\lstinline|f|'' are conventional names for objects and arrows, respectively.
